---
title: "Reverse Engineering Project"
author: "Student names here"
date: "Date here"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

In this notebook, we are reverse engineering the story,As police struggle to solve homicides, Baltimore residents see an ‘open season for killing’, https://www.washingtonpost.com/investigations/as-police-struggle-to-solve-homicides-baltimore-residents-see-an-open-season-for-killing/2018/12/26/7ee561e4-fb24-11e8-8c9a-860ce2a8148f_story.html

## Load libraries

Loading required libraries for this analysis.

```{r echo=FALSE, message=FALSE}
library(tidyverse)
library(lubridate)
library(janitor)
```

## Load and Cleaning Data

In this section, describe the source of the data, write a basic data dictionary for data you are working with, and discuss any caveats or issues you discovered working with this data. 

```{r}
# Load required data
homicide_data <-read_csv("data/homicide-data.csv")
# Path to data should be loaded from folder "data" i.e. read_csv("data/name_of_data.csv")

# Clean required data and prepare for analysis if needed. 
homicide_data <- homicide_data |>
  mutate(fixed_date = ymd(reported_date),
         year = year(fixed_date))
```

## Sentences to Engineer
In this notebook, we are reverse engineering five sentences from the story.

### Sentence 1

* **Sentence text**: As Baltimore has seen a stunning surge of violence, with nearly a killing each day for the past three years in a city of 600,000, homicide arrests have plummeted.
* **Analysis summary**: We found that from 2015-2017 there were 320-342 homicides per year making it around a homicide a day. We filtered the cities just to Baltimore and than used OR code to compile the years the sentence is referring to.

```{r}
# Put code to reverse engineer sentence here
homicide_data <- homicide_data |>
  mutate(reported_date = ymd(reported_date)) |>
  mutate(year = year(reported_date))
  
homicides_per_day <- homicide_data |>
  filter(city == "Baltimore") |>
  filter(year == "2015" | year == "2016" | year == "2017")|>
  group_by(year) |>
  summarize(
    count = n(),
    average_per_day = count/365
  )

# Display results of code below this codeblock
homicides_per_day
```

### Sentence 2

* **Sentence text**: Of 50 of the nation’s largest cities, Baltimore is one of 34 where police now make homicide arrests less often than in 2014, according to a Washington Post analysis.
* **Analysis summary**: We able to find 24 cities that had less arrest, however, Baltimore was not a city found to have less arrests since 2014. This might've been because data was entered incorrectly or in a different way than we filtered but we couldn't find any apparent issues. 

```{r}
# Put code to reverse engineer sentence here
arrests_2014 <- homicide_data |>
  filter(year == "2014" & disposition == "Closed by arrest") |>
  group_by(city) |>
  summarize(arrests_2014 = n())

arrests_2017 <- homicide_data |>
  filter(year == "2017" & disposition == "Closed by arrest") |>
  group_by(city) |>
  summarize(arrests_2017 = n())

homicide_arrests<- arrests_2014 |>
  left_join(arrests_2017) |>
  mutate(pct_change = ((arrests_2017 - arrests_2014)/arrests_2014)*100) |>
  filter(pct_change < 0)

# Display results of code below this codeblock
homicide_arrests
```

### Sentence 3

* **Sentence text**: Baltimore is also one of 30 cities that have seen an increase in homicides in recent years, with the greatest raw number increase in killings of any city other than Chicago, which has four times the population.
* **Analysis summary**: We were able to find that Baltimore was one of 23 cities that had an increase in homicides. We are not sure why there are only 23 in the data and not 30 like in the article. We assume there was some data issue but couldn't find it.

```{r}
# Put code to reverse engineer sentence here
homicide_count_2012 <- homicide_data |>
  filter(year == "2012") |>
  group_by(city)|>
  summarize(homicide_count_2012 = n())

homicide_count_2017 <- homicide_data |>
  filter(year == "2017") |>
  group_by(city) |>
  summarize(homicide_count_2017 = n())

homicide_data_combined <- homicide_count_2017 |>
  left_join(homicide_count_2012) |>
  filter(city != "New York") |>
  mutate(raw_number_increase = homicide_count_2017 - homicide_count_2012) |>
  arrange(desc(raw_number_increase)) |>
  filter(raw_number_increase > 0)

# Display results of code below this codeblock
homicide_data_combined
```

### Sentence 4

* **Sentence text**: For most of the decade before 2015, Baltimore’s annual homicide arrest rate hovered at about 40 percent. Since 2015, the arrest rate hasn’t topped 30 percent in any year.
* **Analysis summary**: We were able to join arrests and homicides and then calculate the rate of which arrests were made for arrests. We were able to confirm the story's findings and also figure out that the number of homicides increased but the arrests didn't.

```{r}
# Put code to reverse engineer sentence here
baltimore_arrests_by_year <- homicide_data |>
  filter(city == "Baltimore", disposition == "Closed by arrest") |>
  group_by(year) |>
  summarize(arrests = n())

baltimore_homicides_by_year <- homicide_data |>
  filter(city == "Baltimore") |>
  group_by(year) |>
  summarize(homicides = n())

baltimore_arrest_rate <- baltimore_arrests_by_year |>
  left_join(baltimore_homicides_by_year, by = "year") |>
  mutate(arrest_rate = (arrests/homicides)*100)

# Display results of code below this codeblock

baltimore_arrest_rate
```

### Sentence 5

* **Sentence text**: 
* **Analysis summary**: [Write up two to three sentences describing the results of your analysis.  Were you able to confirm the finding? If not, why not?]

```{r}
# Put code to reverse engineer sentence here



# Display results of code below this codeblock

```

-30-