---
title: "pre_lab_08.Rmd"
author: "Derek Willis"
date: "2023-03-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Chapter 26

Visualizing data is becoming a much greater part of journalism. Large news organizations are creating graphics desks that create complex visuals with data to inform the public about important events.

To do it well is a course on its own. And not every story needs a feat of programming and art. Sometimes, you can help yourself and your story by just creating a quick chart, which helps you see patterns in the data that wouldn't otherwise surface.

Good news: one of the best libraries for visualizing data is in the tidyverse and it's pretty simple to make simple charts quickly with just a little bit of code. It's called [ggplot2](https://ggplot2.tidyverse.org/).

Let's revisit some data we've used in the past and turn it into charts. First, let's load libraries. When we load the tidyverse, we get ggplot2.

### Task 1: Load packages

**Task** Run the following code to load packages.

```{r}
library(tidyverse)
library(lubridate)
```

The dataset we'll use is 911 overdose calls from Baltimore County. Let's load it.

### Task 2: Load data

**Task** Run the following code to load the data. In the space below, indicate how many rows and columns it has. **Answer** 3822 rows and 4 columns

```{r}
baltco_911_calls <- read_csv("data/baltco_911_calls.csv")
```

## Bar charts

The first kind of chart we'll create is a simple bar chart.

It's a chart designed to show differences between things -- the magnitude of one thing, compared to the next thing, and the next, and the next.

So if we have thing, like a county, or a month, or a day of the week, and then a count of that group, we can make a bar chart.

So what does the chart of the top months with the most 911 overdose calls look like?

First, we'll add a month column to our dataframe using lubridate.

### Task 3: Add a month column to our 911 calls dataframe using lubridate and calculate the number of calls for each month

**Task** Run the following code to add a month column to our dataframe using lubridate and count the number of calls for each month. In the space below, explain what each step of the code below is doing in English. Note the "label=TRUE" part in the mutate statement. **Answer** First, you are taking the baltco_911_calls dataset and renaming it to baltco_911_calls_month. Second, the mutate makes a new column that takes the date column and telling it to just extract the month and then instead of say January being 1, its Jan. Third, we ask R to just look at the new month column and count the amount of times a month shows up in the data. Fourth, we tell R to sort the data from the month with the most 911 calls at the top and the least at the bottom. Finally, we just enter baltco_911_calls_month to make the data show up below the code. 

```{r}
baltco_911_calls_by_month <- baltco_911_calls |>
  mutate(month = month(date, label=TRUE)) |> 
  group_by(month) |> 
  summarize(total_calls = n()) |> 
  arrange(desc(total_calls))

baltco_911_calls_by_month
```

Now let's create a bar chart using ggplot.

With ggplot, the first thing we'll always do is draw a blank canvas that will contain our chart. We start with our dataframe name, and then (\|\>) we invoke the ggplot() function to make that blank canvas. All this does is make a gray box, the blank canvas that will hold our chart.

### Task 4: Create an empty ggplot container

**Task** Run the following code. Explain what you see after running the code in the space below. **Answer** Just a gray square appeared.

```{r}
baltco_911_calls_by_month |>
  ggplot() +
  geom_bar(aes(x = month, weight = total_calls))
```

Next we need to tell ggplot what kind of chart to make.

In ggplot, we work with two key concepts called geometries (abbreviated frequently as geom) and aesthetics (abbreviated as aes).

Geometries are the shape that the data will take; think of line charts, bar charts, scatterplots, histograms, pie charts and other common graphics forms.

Aesthetics help ggplot know what component of our data to visualize -- why we'll visualize values from one column instead of another.

In a bar chart, we first pass in the data to the geometry, then set the aesthetic.

In the codeblock below, we've added a new function, geom_bar().

Using geom_bar() -- as opposed to geom_line() -- says we're making a bar chart.

Inside of that function, the aesthetic, aes, says which columns to use in drawing the chart.

We're setting the values on the x axis (horizontal) to be the abbreviation of the month. We set weight to total calls, and it uses that value to "weight" or set the height of each bar.

One quirk here with ggplot.

After we've invoked the ggplot() function, you'll notice we're using a + symbol. It means the same thing as \|\> -- "and then do this". It's just a quirk of ggplot() that after you invoke the ggplot() function, you use + instead of \|\>. It makes no sense to me either, just something to live with.

### Task 5: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below. **Answer** A bar chart with the months going left to right.

```{r}
baltco_911_calls_by_month |>
  ggplot() +
  geom_bar(aes(x=month, weight=total_calls))
```

This is a very basic chart. But it's hard to derive much meaning from this chart, because the months aren't ordered from highest to lowest by total_calls We can fix that by using the reorder() function to do just that:

### Task 6: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below. What has changed from the last codeblock? **Answer** Now, the chart is sorting the chart with the months with the least calls on the left going to most to the right. The code includes "reorder" which is why it changed.

```{r}
baltco_911_calls_by_month |>
  ggplot() +
  geom_bar(aes(x=reorder(month,total_calls), weight=total_calls))
```

This is a little more useful. But the bottom is kind of a mess, with overlapping names. We can fix that by flipping it from a vertical bar chart (also called a column chart) to a horizontal one. coord_flip() does that for you.

### Task 7: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below. What has changed from the last codeblock? **Answer** The chart is now coming from the left side and its because we used "coord_flip()" to flip the chart. 

```{r}
baltco_911_calls_by_month |>
  ggplot() +
  geom_bar(aes(x=reorder(month,total_calls), weight=total_calls)) +
  coord_flip()
```

Is this art? No. Does it quickly tell you something meaningful? It does.

We're mainly going to use these charts to help us in reporting, so style isn't that important.

But it's worth mentioning that we can pretty up these charts for publication, if we wanted to, with some more code. To style the chart, we can change or even modify the "theme", a kind of skin that makes the chart look better.

It's kind of like applying CSS to html. Here I'm changing the theme slightly to remove the gray background with one of ggplot's built in themes, theme_minimal().

### Task 8: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below. What has changed from the last codeblock? **Answer** Now the background and the lines are less apparent because we did the "theme_minimal" function to change the theme to be more minimalistic. 

```{r}
baltco_911_calls_by_month |>
  ggplot() +
  geom_bar(aes(x=reorder(month,total_calls), weight=total_calls)) +
  coord_flip() + 
  theme_minimal()
```

The ggplot universe is pretty big, and lots of people have made and released cool themes for you to use. Want to make your graphics look kind of like [The Economist's](https://www.economist.com/) graphics? There's a theme for that.

First, you have to install and load a package that contains lots of extra themes, called [ggthemes](https://yutannihilation.github.io/allYourFigureAreBelongToUs/ggthemes/).

### Task 9: Run the following code

**Task** Run the following code.

```{r}
install.packages('ggthemes')
library(ggthemes)
```

And now we'll apply the economist theme from that package with theme_economist()

### Task 10: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below. What has changed from the last version of the chart? **Answer** Now instead of a minimalist look, the background is blue and the lines are more apparent because we used "theme_economist"

```{r}
baltco_911_calls_by_month |>
  ggplot() +
  geom_bar(aes(x=reorder(month,total_calls), weight=total_calls)) +
  coord_flip() + 
  theme_economist()
```

Those axis titles are kind of a mess. Let's change "count" on the x axis to "total calls" and change "reorder(month,total_calls)" to "month". And while we're at it, let's add a basic title and a source as a caption. We'll use a new function, labs(), which is short for labels.

### Task 11: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below. What has changed from the last codeblock? **Answer** Now there is a header at the top and x and y axis are labeled as month and total calls instead of reorder(month, total calls) and count respectively. Also a source is included in the bottom right. This was all done using the "labs" function and using title, x, y, and caption variables to include the changes. 

```{r}
baltco_911_calls_by_month |>
  ggplot() +
  geom_bar(aes(x=reorder(month,total_calls), weight=total_calls)) +
  coord_flip() + 
  theme_economist() +
  labs(
    title="More 911 Overdose Calls in Warmer Months",
    x = "month",
    y = "total calls",
    caption = "source: Baltimore County"
    
  )
```

Viola. Not super pretty, but good enough to show an editor to help them understand the conclusions you reached with your data analysis.

## Line charts

Let's look at how to make another common chart type that will help you understand patterns in your data.

Line charts can show change over time. It works much the same as a bar chart, code wise, but instead of a weight, it uses a y.

So, let's use the same Baltimore County data and generate a count of contributions for each date in our data.

### Task 12: Create a dataframe of 911 calls by date

**Task** Run the following code. Explain what you see after running the code in the space below. **Answer** Instead of just the months, its now the entire date with the certain numbers of calls per day. 

```{r}
baltco_911_calls_by_date <- baltco_911_calls |>
  group_by(date) |>
  summarise(
    total_calls=n()
  )

baltco_911_calls_by_date 
```

And now let's make a line chart to look for patterns in this data.

We'll put the date on the x axis and total contributions on the y axis.

### Task 13: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below. **Answer** I see a line graph but it's jagged and hard to interpret.

```{r}
baltco_911_calls_by_date |>
  ggplot() + 
  geom_line(aes(x=date, y=total_calls))
```

It's not super pretty, but there's a bit of a pattern here: the number of calls fluctuates between 5 and 20 a day for most of this period, and then jumps way up at certain points during the year. In particular there are spikes in July, early October and January.

Right now, it's kind of hard to see specifics, though. When did some of those smaller spikes and troughs happen?

We can't really tell. So let's modify the x axis to have one tick mark and label per month. We can do that with a function called scale_x\_date().

We'll set the date_breaks to appear for every week; if we wanted every month, we'd say date_breaks = "1 month". We can set the date to appear as month abbreviated name (%b) and day (%d).

### Task 14: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below. **Answer** The graph looks the same but now the x axis has a point every week and it's impossible to see.

```{r}
baltco_911_calls_by_date |>
  ggplot() + 
  geom_line(aes(x=date, y=total_calls)) + 
  scale_x_date(date_breaks = "1 week", date_labels = "%b %d")
```

Those are a little hard to read, so we can turn them 45 degrees to remove the overlap using the theme() function for styling. With "axis.text.x = element_text(angle = 45, hjust=1)" we're saying, turn the date labels 45 degrees.

### Task 15: Run the following code

**Task** Run the following code. Explain what you see after running the code in the space below. **Answer** Now it is easier to see the x axis values because they are at an angle. 

```{r}
baltco_911_calls_by_date |>
  ggplot() + 
  geom_line(aes(x=date, y=total_calls)) + 
  scale_x_date(date_breaks = "1 week", date_labels = "%b %d") +
  theme(
    axis.text.x = element_text(angle = 45,  hjust=1)
  )
```

Again, this isn't as pretty as we could make it. But by charting this, we can quickly see some outliers that could guide our reporting.

We're just scratching the surface of what ggplot can do, and chart types. There's so much more you can do, so many other chart types you can make. But the basics we've shown here will get you started.
